Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms

### 2020-03-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Comments and changes to research report manuscript to fit new simulation (until methods section)

Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, \S~6.5.2]. Therefore diagnostic evaluation of convergence would be preferred. However, not known whether MCMC diagnostics are appropriate for MICE ... The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? Defaults in: SPSS (10; https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus (100??; https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata (10; https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia (NA, because resampling, not convergence; https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI (30; https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

## Terminology

<!-- The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem.  -->
This note follows notation and conventions of `mice` [@mice].
<!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3].  -->
<!-- Basic familiarity with MI methodology is assumed.   -->

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as 'incomplete' or 'observed' data. The missing part of $Y$ is denoted by $Y_{mis}$. Which parts of $Y$ are missing is determined by the 'missingness mechanism'.

Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
This note focuses on the algorithmic properties of the imputation step.

The MI algorithm in `mice` has an iterative nature. For each missing datapoint in $Y_{mis}$, $m$ 'chains' of potential values are sampled. Only the ultimate sample that each chain lands on is imputed. The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.  This is repeated $m$ times. Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $Y_{mis}$, . Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. , and subsequently used in the analysis and pooling steps. 
The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.

@@

In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).

@@

The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. With an adapted version of $\widehat{R}$, proposed by @veht19, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p.~898].

The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. The stationarity component may be assessed through autocorrelations. An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity. Other convergence diagnostics are outside of the scope of this study.

All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore!**).

To define $\widehat{R}$, we follow notation by [@veht19, p.~5]. Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.~147]. In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, \S~2.5.2]. That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence. We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study.

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p~437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.

@@

We use model-based simulation in `R` to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage). These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).

@@

### Data Simulation

The point of origin for all simulation conditions is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:

@@

The quantity of scientific interest is regression coefficient $\beta_1$. The data generating model is a multivariate normal distribution with the following means structure and variance-covariance matrix:

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot].
\end{align*}

Data is simulated with the function `mvtnorm::rmvnorm()`, and outcome variable $Y$ is subsequently calculated as 
$$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ [@rubin87]. The complete data is 'amputed' once for each simulation repetition. That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed. %The resulting amputed data is equal for all simulation conditions in the same repetition.

@@

From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).  
We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset. We report the maximum (absolute) value as the 'worst linear predictor' of convergence. $\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations. $AC$ is computed with function `stats::acf()`.

@@

### Convergence Diagnostics

@@

### Simulation Diagnostics

>>>>

Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Therefore diagnostic evaluation of convergence would be preferred. However, not known whether MCMC diagnostics are appropriate for MICE ... The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 1; [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

## Notation

<!-- The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem.  -->
<!-- This note follows notation and conventions of `mice` [@mice]. -->
<!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3].  -->
<!-- Basic familiarity with MI methodology is assumed.   -->

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$). The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.

<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

<!-- The result is $m$ completed datasets, consisting of $y_{obs}$ and   The MI algorithm in `mice` has an iterative nature. For each missing datapoint in $y_{mis}$, $m$ 'chains' of potential values are sampled. -->
<!-- Which parts of $y$ are missing is determined by the 'missingness mechanism'. -->

<!-- Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$. -->
<!-- This note focuses on the algorithmic properties of the imputation step.  -->

<!-- Only the ultimate sample that each chain lands on is imputed. The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.  -->

<!-- This is repeated $m$ times. Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$, . Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$. The chains are terminated after a predefined number of iterations. , and subsequently used in the analysis and pooling steps.  -->
<!-- The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.  -->

@@

In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling). (**add example/guidelines visual inspection here?**)

@@

The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. With an adapted version of $\widehat{R}$, proposed by @veht19, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p. 898].

<!-- The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. The stationarity component may be assessed through autocorrelations. An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity. Other convergence diagnostics are outside of the scope of this study. -->

Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore!**).

## Potential scale reduction factor

To define $\widehat{R}$, we follow notation by [@veht19, p. 5]. Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

## Autocorrelation

Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p. 147]. In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence. We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study.

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.

@@

The convergence of `mice` is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R]. The simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).

<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->

<!-- to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage). These diagnostics are aggregated across repetitions to obtain their average value per simulation condition. -->

@@

A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:

@@

The data generating model is a multivariate normal distribution

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot]
\end{align*}

and simulated with the function `mvtnorm::rmvnorm()`. Outcome variable $Y$ is subsequently calculated as 
$$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->

The complete data is 'amputed' once for each simulation repetition. That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data. **(change this: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). **

@@

**(tot hier doorgelezen)**

<!-- From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).   -->
<!-- We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset. We report the maximum (absolute) value as the 'worst linear predictor' of convergence.  -->
$\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations. $AC$ is computed with function `stats::acf()`.

@@

```{r}
# plot results
load("../Results/results.Rdata")

```

## Convergence Diagnostics

@@

## Simulation Diagnostics
